Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-26 papers/paper/ms.tex
update ms

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
    }

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages.  Opaque lines
  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
  evolutionary state; transparent lines do not.  The bottom row shows
  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

\subsection{A Paucity Of Young Stars And Planets}

@@

The similarity of the KOI host and Kepler star age distributions adds
some nuance to the argument that young transiting planets are hard to
detect due to the photometric variability of their host stars.  If
true, this statement seems to hold only for the very youngest
($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
KOI host age distribution relative to the parent stellar sample.
Examining scatter plots of the planet properties as a function of
age, we similarly find that fewer systems at $t_{\rm
gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
than at $t_{\rm gyro}$$>$0.4\,Gyr.

An important separate selection effect concerns Kepler's ability to
detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
issue, and found for Sun-like stars that the fraction of rotation
signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
that it drops to almost zero by $\approx$5\,Gyr.  This trend is
opposite in functional form to our derived age distributions\replaced{.}{and
	most of the stars in our sample are indeed G dwarfs (see e.g.,~the $t_{\rm gyro}$--$T_{\rm eff}$ distributions 
	in Appendix~\ref{app:age_diagnostic}).}

@@

@@

In the entire planet sample (i.e.~without imposing any quality cuts)
this yielded \allagesyesconsistent\ consistent cases,
\allagesmaybeconsistent\ maybe consistent cases, and
\allagesnoconsistent\ inconsistent cases.  Rephrased, in the overall
sample, \fracconsistentallages\% of stars have consistent rotation and
lithium-based ages (\fracpotentiallyconsistentallages\% potentially
consistent).

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
more interesting of the two is Kepler-786, an early K dwarf with
$t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
rotation period that implies $t_{\rm
gyro}$$\approx$4.4,Gyr.  Other age indicators in the
spectrum similarly suggest that the star is not young.  It 
therefore appears to be anomalously lithium-rich.

@@

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than stellar age and mass can
influence stellar rotation rates.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
is that our rotation-based ages used photometric effective
temperatures (Section~\ref{sec:stellarprops}), even though
spectroscopic temperatures are available for the planet-hosting
subset.  This decision was driven by a desire for homogeneity 
irrespective of planet-hosting status.  However it implies that the
estimated planet ages might shift if one were to account for the
spectroscopic information.

@@

A broader question, beyond the scope of this work, concerns the ages
of stars $\gtrsim$4\,Gyr old.  For age-dating models based only on
rotation rates, an important challenge is that the detectability of
the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
is at the limits of the Kepler data \citep{2022ApJ...937...94M}.  In
the ``old star'' regime, methods that leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both the
evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most capable of
providing useful age constraints for single stars over the full
span of their main-sequence lifetimes.
%Incorporating
%kinematics \citep{2021AJ....161..189L} seems likely to remain most
%useful for stellar ensembles, due to the stochasticity of kinematic
%heating.

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.  Some care is required since
beyond age, exoplanet demographics also depend on stellar metallicity
and mass \citep[e.g.][]{Petigura_2018,Miyazaki2023}.

>>>>

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
    }

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
    }
%
%	\vspace{-0.35cm}
%	\subfloat{
%		\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
%	}
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages.  Opaque lines
  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
  evolutionary state; transparent lines do not\added{ (see
  Section~\ref{subsec:flags})}.  The \replaced{bottom}{middle} row shows
  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  
  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

\subsection{Rotation-Based Age Distribution}

\subsubsection{A Paucity Of Young Stars And Planets}

@@

The similarity of the KOI host and Kepler star age distributions adds
some nuance to the argument that young transiting planets are hard to
detect due to the photometric variability of their host stars.  If
true, this statement \replaced{seems to}{might} hold only for the very youngest
($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
KOI host age distribution relative to the parent stellar sample.
Examining scatter plots of the planet properties as a function of
age, we similarly find that fewer systems at $t_{\rm
gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
than at $t_{\rm gyro}$$>$0.4\,Gyr.

@@

\subsubsection{Nuances in the Age Distribution}

\added{{\it Completeness}---Figure~\ref{fig:hist_tgyro} has a gray
overlay beyond $\gtrsim$3\,Gyr because Kepler's sensitivity to
rotation signals diminishes at old ages.} \citet{2022ApJ...937...94M}
studied \deleted{this issue}\added{Kepler's completeness to rotation
signals using MIST-derived isochrone ages}, and found for Sun-like
stars that the fraction of rotation signals \added{that are
}detectable \deleted{by Kepler }is near unity up to $\approx$3\,Gyr,
and that it drops to almost zero by $\approx$5\,Gyr.  \deleted{This trend is
opposite in functional form to our derived age
distributions.}\added{This estimate suggests that while aspects of
Figure~\ref{fig:hist_tgyro} might be interpretable in terms of the Galaxy's star
formation history at $\lesssim$3\,Gyr, at older ages incompleteness
likely precludes any such interpretation.}

\added{{\it Temperature sensitivity}---There are quantitative differences
in Figure~\ref{fig:hist_tgyro} between the
K dwarfs (4400--5400\,K) and the FGK stars (3800--6200\,K).  For instance,
the distribution peaks at 2.3\,Gyr for K dwarfs, and at 2.9\,Gyr for
the FGK sample.  The peak is also sharper for the K dwarfs;
its decrease from 2.3-3\,Gyr is in an age range in which we believe the
rotation period catalog completeness to be near unity.  We
discuss further details of the $t_{\rm gyro}$--$T_{\rm eff}$
distributions in Appendix~\ref{app:age_diagnostic}, and propose possible
explanations for these differences in Section~\ref{subsec:caveats}. }

@@

In the entire planet sample (i.e.,~without imposing any quality cuts)
this yielded \allagesyesconsistent\ consistent cases,
\allagesmaybeconsistent\ maybe consistent cases, and
\allagesnoconsistent\ inconsistent cases.  Rephrased, in the overall
sample, \fracconsistentallages\% of stars have consistent rotation and
lithium-based ages (\fracpotentiallyconsistentallages\% potentially
consistent).

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
more interesting of the two is Kepler-786, an early K dwarf with
$t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
rotation period that implies $t_{\rm
gyro}$$\approx$4.4\,Gyr.  Other age indicators in the
spectrum similarly suggest that the star is not young.  It 
therefore appears to be anomalously lithium-rich.

@@

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than \deleted{stellar }age and mass can
influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
metallicity, differential rotation, and tidal spin-up, while observational
challenges include the difficulty of measuring small-amplitude,
long-period signals that are only quasi-coherent due to the starspot
lifetime approaching the stellar rotation period.  In the following, we attempt to
quantify the importance of these effects in our analysis, ordered by
what we perceive to be their descending importance.}

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
%		\subfloat{
%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
%		}
%		
%		\vspace{-0.35cm}
%		\subfloat{
%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
%		}
		%
		%	\vspace{-0.35cm}
			\subfloat{
				\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
			}
	\end{center}
	\vspace{-0.6cm}
	\caption{{\bf Variant of Figure~\ref{fig:hist_tgyro}}, with rotation-derived age distributions that
			correct for the estimated (estm.) effect of unresolved $\lesssim$0.5\,AU binaries. 
		Even after applying our quality cuts (Section~\ref{subsec:flags}), up to
		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
		$P_{\rm rot}$$<$10\,days could be unresolved binaries.
		These distributions correct for this possibility by randomly removing
		what we expect to be the maximum possible number of such systems; the actual
		correction relative to Figure~\ref{fig:hist_tgyro} might be smaller.
		\label{fig:hist_tgyro_corr}
	}
\end{figure*}

{\it Binaries}:
\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
are biased toward rapid rotation \citep[e.g.][and many studies
thereafter]{Meibom_2007}.}{Binaries are often biased
toward rapid rotation, even at intermediate separations of $\lesssim$10\,AU \citep[e.g.][]{Meibom_2007,Gruner_2023}.}
\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
Kepler stars with rotation periods between 1.5 and 7\,days lie
0.3\,mag above the main sequence, compared with 28\% of their full
rotation sample.
We attempted to mitigate the effects of binarity with five distinct
quality flags (Section~\ref{subsec:flags}), and this attempt was at least
somewhat effective, based on the peak near zero in the raw
distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
cleaned distributions.
However, the sensitivity of our quality flags is hard to assess.  One specific concern is that these
quality flags might have minimal sensitivity to binaries with
separations of 0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
orbital periods of astrometric binaries in Gaia DR3 \citep{GaiaCollaboration2023}.  }

\added{ If we were not at all sensitive to
binaries with separations of 0.05-0.5\,AU, then based on
the binary fraction and separation distribution from
\citet{Raghavan2010}, as much as $\approx$5\% of our ``gyro
applicable'' sample
(\fivepctnuniqstarsantosrotgyroappl/\nuniqstarsantosrotgyroappl\
stars) might be contaminated.  In a worst case scenario, all of
these unidentified binaries might be able to induce $P_{\rm rot}$$<$10
days, which would then be mixed in with the bona fide young, single stars.
In our actual sample of apparently clean stars,
\ngyroappllttendays/\nuniqstarsantosrotgyroappl\ had $P_{\rm
rot}$$<$10\,days.

Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
$P_{\rm rot}$$<$10\,day stars.
There are three important aspects of this result, relative to the original
distributions in Figure~\ref{fig:hist_tgyro}.
The first is that it would imply a much larger drop in the star formation rate,
by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr,
over the past three gigayears.
The second is that it allows for a constant rate of decline in star formation
since $\approx$2.8\,Gyr ago, rather than having a ``kink'' at 1.5\,Gyr,
as in Figure~\ref{fig:hist_tgyro}.
And finally, these corrected distributions show closer agreement with SFHs
derived from CMD fitting.
We caution however that this correction assumes that {\it none} of our
quality flags were able to identify any of these unresolved binaries.
Since we did in fact include flags based on e.g.~location in the HR diagram,
this correction is best interpreted as an upper limit on the possible impact of binarity.}

%TODO TODO TODO
\added{{\it Erroneous period measurements:} }

\added{
{\it Metallicity:}
}

\added{
{\it Differential rotation:}
}

%TODO TODO TODO
%TODO TODO TODO
%TODO TODO TODO

Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
is that our rotation-based ages used photometric effective
temperatures (Section~\ref{sec:stellarprops}), even though
spectroscopic temperatures are available for the planet-hosting
subset.  This decision was driven by a desire for homogeneity 
irrespective of planet-hosting status.  However it implies that the
estimated planet ages might shift if one were to account for the
spectroscopic information.

@@

%Incorporating
%kinematics \citep{2021AJ....161..189L} seems likely to remain most
%useful for stellar ensembles, due to the stochasticity of kinematic
%heating.

@@

\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
old.}\added{This work was limited to stars younger than 4\,Gyr}. For
age-dating models based only on rotation rates, an important challenge
is that the detectability of the rotation signals for 3-10\,Gyr stars,
especially Sun-like stars, is at the limits of the Kepler data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
the evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
capable of providing useful age constraints for single stars over the
full span of their main-sequence lifetimes.

{\it \added{Planet} occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.  Some care is required since
beyond age, exoplanet demographics also depend on stellar metallicity
and mass \citep[e.g.][]{Petigura_2018,Miyazaki2023}.
